Parsing Postfix log files

### 2008-08-25 doc/logparser-acronyms.tex
Changes made following suggestions from diction.

\newacronym{API}{Application Programming Interface}{description={
    One of the fundamental concepts when writing programs is the reuse of
    existing code, so that each program does not reinvent the wheel.  When
    a body of code is intended to be reused repeatedly, the user of this
    code needs to be informed of the functionality provided by the code.
    An API defines the interface provided to the user, and acts as a
    contract between the user and the provider: if the user adheres to the
    API the provider guarantees it will work, while the provider is free to
    change the implementation as long as the API is preserved.
}}

\newacronym{SMTP}{Simple Mail Transfer Protocol}{description={
    SMTP is the protocol which transfers mail from the sender to the
    recipient across the Internet.  It is a simple, human readable, plain
    text protocol, making it quite easy to test and debug problems with it.
    A detailed description of \SMTP{} is beyond the scope of this document:
    the original protocol definition is \RFC{}~821~\cite{RFC821}, updated
    in \RFC{}~2821~\cite{RFC2821}; introductory guides can be found
    at~\cite{smtp-intro-01,smtp-intro-02}.
}}

\newacronym{LMTP}{Local Mail Transfer Protocol}{description={
    LMTP is a protocol derived from SMTP that removes the need for the
    server to maintain a mail delivery queue, instead relying on the client
    to maintain it.  Typically the client would be an MTA, and the server
    would be a delivery agent or a mail store.  Full details are available
    in~\cite{lmtp-rfc}.
}}

\newacronym{MTA}{Mail Transfer Agent}{description={
    A Mail Transfer Agent sends and/or receives mail via
    SMTP\backslashat{}.
}}

@@

\newacronym{ISP}{Internet Service Provider}{description={
    An ISP is a company which sells Internet access to consumers.
}}

\newacronym{LMA}{Log Mail Analyzer}{description={
    One of the other Postfix log parsers reviewed and the only example of
    published prior art~\cite{log-mail-analyser} found by the author.
}}

@@

\newacronym{pid}{Process Identifier}{description={
    There may be multiple copies of any program executing at any one time,
    so the program's name is not suitable as a distinguishing identifier;
    instead each process executing is given a pid which is guaranteed to be
    unique for the lifetime of the process.  Once the process has
    completed, the pid may be reused, as they are drawn from a finite pool.
}}

\newacronym{UCE}{Unsolicited Commercial Email}{description={
    UCE is a more restrictive definition of spam than most people would
    use: it only covers mail that is explicitly commercial, thus excluding
    viruses, Bayesian poisoning mails, backscatter, and those annoying
    chain letters you get from friends.
}}

@@

\newacronym{LCD}{Lowest Common Denominator}{description={
    LCD is a mathematical term which is used figuratively to mean the least
    useful member of a set of alternatives.
}}

@@

\newacronym{NLP}{Natural Language Processing}{description={
    Natural Language Processing~\cite{Wikipedia-nlp} attempts to increase
    our understanding of the languages normally used by humans (e.g.\
    English, Spanish, Japanese), with the goal of writing programs which
    can understand human languages.
}}

>>>>

\newacronym{API}{Application Programming Interface}{description={
    One of the fundamental concepts when writing programs is the reuse of
    existing code, so that each program does not reinvent the wheel.  When
    a body of code is intended to be reused repeatedly, the user of this
    code needs to be informed of the functionality provided by the code.
    An API defines the interface provided to the user, and acts as a
    contract between the user and the provider: if the user adheres to the
    API the provider guarantees it will work, but is free to change the
    implementation provided the API is preserved.
}}

\newacronym{SMTP}{Simple Mail Transfer Protocol}{description={
    SMTP is the protocol that transfers mail from the sender to the
    recipient across the Internet.  It is a simple, human readable, plain
    text protocol, making it quite simple to test and debug problems with
    it.  A detailed description of \SMTP{} is beyond the scope of this
    document: the original protocol definition is \RFC{}~821~\cite{RFC821},
    updated in \RFC{}~2821~\cite{RFC2821}; introductory guides can be found
    at~\cite{smtp-intro-01,smtp-intro-02}.
}}

\newacronym{LMTP}{Local Mail Transfer Protocol}{description={
    LMTP is a protocol derived from SMTP that removes the need for the
    server to maintain a mail delivery queue, instead relying on the client
    to maintain it.  Typically the client is an MTA, and the server a
    delivery agent or a mail store.  Full details are available
    in~\cite{lmtp-rfc}.
}}

\newacronym{MTA}{Mail Transfer Agent}{description={
    A Mail Transfer Agent sends and receives mail via
    SMTP\backslashat{}.
}}

@@

\newacronym{ISP}{Internet Service Provider}{description={
    An ISP is a company that sells Internet access to consumers.
}}

\newacronym{LMA}{Log Mail Analyzer}{description={
    One of the other Postfix log parsers reviewed and the only example of
    published prior art~\cite{log-mail-analyser} found that specifically
    deals with parsing Postfix logs.
}}

@@

\newacronym{pid}{Process Identifier}{description={
    There may be multiple copies of any program executing at any one time,
    so the program's name is not suitable as a distinguishing identifier;
    instead each process executing is given a pid that is guaranteed to be
    unique for the lifetime of the process.  Once the process has
    completed, the pid may be reused, as they are drawn from a finite pool.
}}

\newacronym{UCE}{Unsolicited Commercial Email}{description={
    UCE is a more restrictive definition of spam than most people use: it
    only covers mail that is explicitly commercial, thus excluding viruses,
    Bayesian poisoning mails, backscatter, and those annoying chain letters
    you get from friends.
}}

@@

\newacronym{LCD}{Lowest Common Denominator}{description={
    LCD is a mathematical term that is used figuratively to mean the least
    useful member of a set of alternatives.
}}

@@

\newacronym{NLP}{Natural Language Processing}{description={
    Natural Language Processing~\cite{Wikipedia-nlp} attempts to increase
    our understanding of the languages normally used by humans (e.g.\
    English, Spanish, Japanese), with the goal of writing programs that
    can understand human languages.
}}

### 2008-08-25 doc/logparser-architecture.tex
Changed normal to optimal.
Changes made following suggestions from diction.

XXX REWRITE AS NECESSARY TO MERGE AND SUPPLEMENT THE EXISTING CONTENT\@.

@@

    \item [actions] Actions are invoked to deal with a log line once it has
        been identified by the rules: actions modify data structures,
        handle complications, and cause data to be saved to the database.
        The actions perform the work of reconstructing the journey a mail
        takes through Postfix.  Details of the actions available in the
        Postfix parser can be found in \sectionref{actions-in-detail}, and
        \sectionref{adding new actions} describes the process of adding new
        actions.

    \item [framework] The framework is responsible for loading rules,
        managing data structures, finding the rule which matches each log
        line, invoking the correct action, etc\@.  The framework is
        described in detail in \sectionref{framework}.  XXX EXTEND
        FRAMEWORK SECTION IF NECESSARY\@.

@@

It may help to think of the rules and actions as components which plug into
the framework.

@@

Decoupling the actions from the framework simplifies both framework and
actions: the framework provides services to the actions, and does not need
to deal with the complications which arise, or the task of reconstructing a
mail's journey through Postfix; actions benefit from having services
provided by the framework, freeing them to concentrate on the task of
accurately reconstructing each mail's journey through Postfix and dealing
with the complications described in \sectionref{additional complications}.

Separating the rules from the actions and framework makes it possible to
parse new log lines without modifying the core parsing algorithm.  Adding a
new rule with the action to invoke and a \regex{} to match the log lines is
trivial in comparison to understanding a program's entire parsing
algorithm, identifying the correct location to change and making the
appropriate changes.  Bear in mind that the changes must be made without
adversely affecting existing parsing, particularly as there may be edge
cases which are not immediately obvious ---
\sectionref{yet-more-aborted-delivery-attempts} describes a complication which
occurs only four times in \numberOFlogFILES{} log files tested.  Requiring
changes to the parser's code also complicates upgrades, as the changes must
be preserved during the upgrade, and may clash with changes made by the
developer (\sectionref{adding new actions} describes safely adding new
actions).  \parsername{} allows the user to add new rules to the database
without changing the parsing algorithm, unless the new log lines to be
parsed require functionality not already provided by the algorithm.  If the
new log lines do require new functionality, new actions can be added to the
parser without modifying existing actions or other parts of the algorithm;
only in the rare case that the new actions require support from other
sections of the code will more extensive changes be required.

There is some similarity between the parser's design and William Wood's
\ATN{}~\cite{atns, nlpip}, a tool used in Computational Linguistics for
creating grammars to parse or generate sentences.  The resemblance between
\ATN{} and the parser is accidental, but it is interesting how two
apparently different approaches share an underlying separation of concerns;
this appears to be a natural division of responsibility and functionality.
\Tableref{Similarities between ATN and this architecture} shows the
similarities.

@@

\begin{table}[ht]
    \caption{Similarities between ATN and this architecture}
    \empty{}\label{Similarities between ATN and this architecture}
    \begin{tabular}[]{lll}
        \tableline{}
        \ATN{}          & Architecture  & Similarity                \\
        \tableline{}
        Networks        & Algorithm     & Determines the sequence 
                                        of transitions              \\
                        &               & or actions which 
                                        constitutes a valid  input. \\
        Transitions     & Actions       & Save data and impose
                                        conditions the              \\
                        &               & input must meet to be
                                        considered valid.           \\
        Abbreviations   & Rules         & Responsible for 
                                        classifying input.          \\
        \tableline{}
    \end{tabular}
\end{table}

@@

Rule have certain characteristics which may help in understanding the
parser:

@@

    \item The first matching rule wins: no further rules are tried against
        that log line, but there is a facility for prioritising the rules
        so that more specific rules can be tried first.

@@

The parser does not try to detect overlapping rules; that
responsibility is left to the author of the rules.  Unintentionally
overlapping rules lead to inconsistent parsing and data extraction because
the order in which rules are tried against each line may change between log
files, and the first matching rule wins.  Overlapping rules are frequently
a requirement, allowing a more specific rule to match some log lines and a
more general rule to match the majority, e.g.\ separating \SMTP{} delivery
to specific sites from \SMTP{} delivery to the rest of the world.  The
algorithm provides a facility for ordering overlapping rules: the priority
field in each rule (defaults to zero).  Negative priorities may be useful
for catchall rules.

@@

    \item Compare the results of parsing using sorted, shuffled and
        reversed rules (described in \sectionref{rule ordering for
        efficiency}).  Parse a number of log files using normal sorting,
        then dump a textual representation of the rules, connections and
        results tables.  Repeat with shuffled and reversed rules, starting
        with a fresh database.  If there are no overlapping rules the
        tables from each run will be identical; differences indicate
        overlapping rules.  Which rules overlap can be determined by
        examining the differences in the tables: each result contains a
        reference to the rule which created it, if the references differ
        between runs the two rules referenced in the differing results
        overlap.  Unfortunately this method cannot prove the absence of
        overlapping rules; it can detect overlapping rules, but only if
        there are log lines in the input files which match more than one
        rule.

@@

The intermingling of log entries from different mails immediately rules out
the possibility of handling each mail in isolation; the parser must be
capable of handling multiple mails in parallel, each potentially at a
different stage in its journey, without any interference between mails ---
except in the minority of cases where intra-mail interference is required.
The best way to implement this is to maintain state information for every
unfinished mail and manipulate the appropriate mail correctly for each log
line encountered.

>>>>

XXX REWRITE TO MERGE AND SUPPLEMENT THE EXISTING CONTENT\@.

@@

    \item [actions] Actions are invoked to deal with a log line once it has
        been identified by the rules: actions modify data structures,
        handle complications, and cause data to be saved to the database.
        The actions do the work of reconstructing the journey a mail takes
        through Postfix.  Details of the actions available in the Postfix
        parser can be found in \sectionref{actions-in-detail}, and
        \sectionref{adding new actions} describes the process of adding new
        actions.

    \item [framework] The framework is responsible for loading rules,
        managing data structures, finding the rule that matches each log
        line, invoking the correct action, etc\@.  The framework is
        described in detail in \sectionref{framework}.  XXX EXTEND
        FRAMEWORK SECTION IF NECESSARY\@.

@@

It may help to think of the rules and actions as components that plug into
the framework.

@@

Decoupling the actions from the framework simplifies both framework and
actions: the framework provides services to the actions, and does not need
to deal with the complications that arise, or the task of reconstructing a
mail's journey through Postfix; actions benefit from having services
provided by the framework, freeing them to concentrate on the task of
accurately reconstructing each mail's journey through Postfix and dealing
with the complications described in \sectionref{additional complications}.

Separating the rules from the actions and framework makes it possible to
parse new log lines without modifying the core parsing algorithm.  Adding a
new rule with the action to invoke and a \regex{} to match the log lines is
trivial in comparison to understanding a program's entire parsing
algorithm, identifying the correct location to change and making the
appropriate changes.  Bear in mind that the changes must be made without
adversely affecting existing parsing, particularly as there may be edge
cases that are not immediately obvious ---
\sectionref{yet-more-aborted-delivery-attempts} describes a complication
that occurs only four times in \numberOFlogFILES{} log files tested.
Requiring changes to the parser's code also complicates upgrades, as the
changes must be preserved during the upgrade, and may clash with changes
made by the developer (\sectionref{adding new actions} describes safely
adding new actions).  \parsername{} allows the user to add new rules to the
database without changing the parsing algorithm, unless the new log lines
to be parsed require functionality not already provided by the algorithm.
If the new log lines do require new functionality, new actions can be added
to the parser without modifying existing actions or other parts of the
algorithm; only in the rare event that the new actions require support from
other sections of the code will more extensive changes be required.

There is some similarity between the parser's design and William Wood's
\ATN{}~\cite{atns, nlpip}, a tool used in Computational Linguistics for
creating grammars to parse or generate sentences.  The resemblance between
\ATN{} and the parser is accidental, but it is obvious that the two
different approaches share an underlying separation of concerns; this
appears to be a natural division of responsibility and functionality.
\Tableref{Similarities between ATN and this architecture} shows the
similarities.

@@

\begin{table}[ht]
    \caption{Similarities between ATN and this architecture}
    \empty{}\label{Similarities between ATN and this architecture}
    \begin{tabular}[]{lll}
        \tableline{}
        \ATN{}          & Architecture  & Similarity                \\
        \tableline{}
        Networks        & Algorithm     & Determines the sequence 
                                        of transitions              \\
                        &               & or actions that 
                                        constitutes a valid  input. \\
        Transitions     & Actions       & Save data and impose
                                        conditions the              \\
                        &               & input must meet to be
                                        accepted as valid.          \\
        Abbreviations   & Rules         & Responsible for 
                                        classifying input.          \\
        \tableline{}
    \end{tabular}
\end{table}

@@

Rule have certain characteristics that may help in understanding the
parser:

@@

    \item The first matching rule wins: no further rules are tried against
        that log line, but there is a mechanism for prioritising the rules
        so that more specific rules can be tried first.

@@

The parser does not try to detect overlapping rules; that responsibility is
left to the author of the rules.  Unintentionally overlapping rules lead to
inconsistent parsing and data extraction because the order in which rules
are tried against each line may change between log files, and the first
matching rule wins.  Overlapping rules are frequently a requirement,
allowing a more specific rule to match some log lines and a more general
rule to match the majority, e.g.\ separating \SMTP{} delivery to specific
sites from \SMTP{} delivery to the rest of the world.  The algorithm
provides a mechanism for ordering overlapping rules: the priority field in
each rule (defaults to zero).  Negative priorities may be useful for
catchall rules.

@@

    \item Compare the results of parsing using sorted, shuffled and
        reversed rules (described in \sectionref{rule ordering for
        efficiency}).  Parse several log files using optimal ordering, then
        dump a textual representation of the rules, connections and results
        tables.  Repeat with shuffled and reversed ordering, starting with
        a fresh database.  If there are no overlapping rules the tables
        from each run will be identical; differences indicate overlapping
        rules.  The rules that overlap can be determined by examining the
        differences in the tables: each result contains a reference to the
        rule which created it, if the references differ between runs the
        two rules referenced in the differing results overlap.
        Unfortunately this method cannot prove the absence of overlapping
        rules; it can detect overlapping rules, but only if there are log
        lines in the input files that match more than one rule.

@@

The intermingling of log entries from different mails immediately rules out
the possibility of handling each mail in isolation; the parser must be able
to handle multiple mails in parallel, each potentially at a different stage
in its journey, without any interference between mails --- except in the
minority of cases where intra-mail interference is required.  The best way
to implement this is to maintain state information for every unfinished
mail and manipulate the appropriate mail correctly for each log line
encountered.
